The importance of measurement uncertainty in ecological management

### 2018-01-27 manuscripts/manuscript.Rmd
WIP

As discussed above, absent the short-cut provided by Reed's theorems -- solving for the optimal management policy in a stochastically fluctuating population -- involves a Markov Decision Process (MDP), commonly referred to in ecological literature by its solution method, Stochastic Dynamic Programming (SDP). Once measurement uncertainty is introduced, the problem becomes a Partially Observed Markov Decision Process (POMDP, not be confused with a partially observed or hidden Markov model, HMM, in which no decision process is involved).  The POMDP problem is posed identically to that of the MDP problem, with the addition of an observation process.  The POMDP problem for fisheries question considered here can be summarized as follows:

@@

The optimization problem is to select the action $a_t$ that will maximize the net present utility over all time. Future utility may be discounted by a factor $\gamma$, so that a value  $V$ in $t$ year is valued at $\gamma^t V$ today.  Numerically, each of these functions are defined over a discrete set of possible states, observations, and actions, and can thus be represented as a collection of matrices or tensors.  As in the case of MDP [@Marescot2013], the POMDP problem can be solved by existing and well-studied algorithms. While not unknown to the conservation literature [@Williams2011], algorithms for POMDP have historically scaled quite poorly, and their application has been restricted to contexts with only a handful of possible states and actions [e.g. @Chades2008; @Chades2011; @Fackler2014; @Fackler2014b].  POMDP algorithms remain a very active area in artificial intellegence, and by adapting recent algorithmic developments [@Kurniawati2008] from that field we are able to find solutions for the considerably more complex problems such as the case of fisheries managment, which requires on the order of 100 states and actions to provide sufficient numerical resolution. Readers interested in the technical details of these algorithms should consult the extensive literature in in POMDP methods. For those interested in exploring or applying these methods, we have developed a preliminary R package [@sarsop-pkg], which can be used to replicate any of the results presented here and explore other variations.

@@

Different attention to prior events fundamentally drives the differences between the POMDP strategy and the other strategies and can resolve the uncertainty paradox.  While the catch quota under both TAC and CE stategies can be completely determined given the most recent observation of the stock size by using the policy curves shown in Figure \ref{policy}, this is not the case for the POMDP approach. This fundamental difference is key to understanding the difference in performance and resolving the paradox of uncertainty.  The POMDP policy cannot be specified by the most recent observation alone. Instead, the POMDP policy depends on all prior observations, not just the most recent.  The reason for this complexity comes from the Markov property.  Observations of the state in the perfectly observed system satisfy the Markov property: once we have measured the current biomass exactly, we cannot get any better estimate of the current stock size by studying older measurements. When measurements are uncertain this is no longer the case: intuitively, by comparing the most recent measurement to previous observations we may be able to infer when any given measurement is unusually high or unusually low.  POMDP formalizes this intutition by capturing the information from all previous observations into a *prior belief*.  This prior belief is updated after every subsequent action and observation in accordance with Bayes Law.  The mechanics of this process are well documented in the extensive literature on POMDPs [e.g. @pomdpstuff; @stuff; @more], but for our purposes it is sufficient to observe how this evolving prior belief serves to continually adjust the POMDP policy.

@@

Using modern algorithms we have been able to crack the nut of measurement uncertainty in the optimal management of marine fisheries and debunk long-standing paradoxes that increased measurement uncertainty should be met with larger harvests.  We have demonstrated that both existing optimal control approaches and existing precautionary approaches result in overexploitation and long term declines of fish stocks, as well as reduced economic value, while in contrast, the POMDP approach is able to deliver both ecological and economic recovery despite uncertainty in measurements.  These results have important implications for both global fisheries and ecological management more generally.  In marine fisheries, evidence of such over-exploitation has already been well documented [@Worm2006; @Worm2009; @Costello2016]. Our results suggest that measurement uncertainty may contribute to this pattern, and that reality of measurement errors may frustrate stock recovery under current best practices such as MSY, TAC, and CE.  More generally, the study and application of strategies such as MSY and CE are in no way limited to the decisions in fisheries quotas, but underpin much of conservation and natural resource management in forests and othe terrestrial ecosystems as well [e.g. @Clark1990].  Our results underscore the realization that measurement uncertainty is not negligible.  Neither optimization approaches which ignore uncertainty (like CE) or heuristic approaches to account for that uncertainty (like TAC) are sufficient for this task.  Rather, adaptive approaches such as POMDP, which make use of previous estimates to continually adjust prior beliefs, are required to avoid negative ecological and economic outcomes.  Nevertheless, it may be tempting to see these results as either trivial or impractical, but a closer inspection reveals they are neither.

@@

Third, are the examples considered here too simple?  The strategies we have compared against underpin both current practice and management aspirations.  Together, MSY and TAC-based management reflects much of modern best practice as it is understood in fisheries management and elsewhere. While real-world application of these approaches frequently involves more complex models, the models we have considered here are the very ones under which these strategies were designed and tested, and there is little reason to believe they would perform better under more complex cases such as age-structured models that do not always meet those assumptions [e.g. @Holden2015].  The relatively poor ecological and economic performance of these common best-practices relative to the alternative approach of POMDP is thus reason enough for concern.  This issue is further underscored by the reality of both long-running ecological detioriation [@Worm2006; @Worm2009] and economic costs of global fisheries.

Fourth and last, is the approach taken here too complex to be feasible? The importance of simple and effective rules of thumb in conservation management has been well documented [e.g. @Possingham]. Sophisticated and computationally intensive approaches such as POMDP may seem improbable at a time when many areas of natural resource management, even simple, rule-of-thumb based methods struggle to take root.  And yet, in this same age we rely on far more complex and computationally intensive algorithms to perform far more trivial tasks such as determining what advertisements we see.  Surely a planet that can afford such complex optimization for how advertising space is allocated on a small screen can one day manage it's natural resources and conservation efforts with such tools?

<!--

@@

-->

>>>>

As discussed above, absent the short-cut provided by Reed's theorems -- solving for the optimal management policy in a stochastically fluctuating population -- involves a Markov Decision Process (MDP), commonly referred to in ecological literature by its solution method, Stochastic Dynamic Programming (SDP). The constant escapement (CE) strategy considered here is the result of such an MDP or SDP analysis.  Once measurement uncertainty is introduced, the problem becomes a Partially Observed Markov Decision Process (POMDP, not be confused with a partially observed or hidden Markov model, HMM, in which no decision process is involved).  The POMDP problem is posed identically to that of the MDP problem, with the addition of an observation process, but cannot be solved using SDP directly.  The POMDP problem for fisheries question considered here can be summarized as follows:

@@

The optimization problem is to select the action $a_t$ that will maximize the net present utility over all time. Future utility may be discounted by a factor $\gamma$, so that a value  $V$ in $t$ year is valued at $\gamma^t V$ today.  Numerically, each of these functions are defined over a discrete set of possible states, observations, and actions, and can thus be represented as a collection of matrices or tensors.  While the solution method for MDP problems, stochastic dynamic programming (SDP) is well known and readily implemented in ecological problems [e.g @Mangel1988; @Marescot2013], the POMDP problem cannot be efficiently solved using SDP. While not unknown to the conservation literature [@Williams2011], algorithms for POMDP have historically scaled quite poorly, and their application has been restricted to contexts with only a handful of possible states and actions [e.g. @Chades2008; @Chades2011; @Fackler2014; @Fackler2014b].  POMDP algorithms remain a very active area in artificial intellegence [. @Smallwood1973; @Sondik1978; @Pineau2003; @Kaelbling1998], and by adapting recent algorithmic developments [@Kurniawati2008] from that field we are able to find solutions for the considerably more complex problems such as the case of fisheries managment, which requires on the order of 100 states and actions to provide sufficient numerical resolution. Readers interested in the technical details of these algorithms should consult the extensive literature in in POMDP methods. For those interested in exploring or applying these methods, we have developed a preliminary R package [@sarsop-pkg], which can be used to replicate any of the results presented here and explore other variations.

@@

Different attention to prior events fundamentally drives the differences between the POMDP strategy and the other strategies and can resolve the uncertainty paradox.  While the catch quota under both TAC and CE stategies can be completely determined given the most recent observation of the stock size by using the policy curves shown in Figure \ref{policy}, this is not the case for the POMDP approach. This fundamental difference is key to understanding the difference in performance and resolving the paradox of uncertainty.  The POMDP policy cannot be specified by the most recent observation alone. Instead, the POMDP policy depends on all prior observations, not just the most recent.  The reason for this complexity comes from the Markov property.  Observations of the state in the perfectly observed system satisfy the Markov property: once we have measured the current biomass exactly, we cannot get any better estimate of the current stock size by studying older measurements. When measurements are uncertain this is no longer the case: intuitively, by comparing the most recent measurement to previous observations we may be able to infer when any given measurement is unusually high or unusually low.  POMDP formalizes this intutition by capturing the information from all previous observations into a *prior belief*.  This prior belief is updated after every subsequent action and observation in accordance with Bayes Law.  The mechanics of this process are well documented in the extensive literature on POMDPs [e.g. @Smallwood1973; @Sondik1978; @Kurniawati2008; @Williams2011], but for our purposes it is sufficient to observe how this evolving prior belief serves to continually adjust the POMDP policy.

@@

Using modern algorithms we have been able to crack the nut of measurement uncertainty in the optimal management of marine fisheries and debunk long-standing paradoxes that increased measurement uncertainty should be met with larger harvests.  We have demonstrated that both existing optimal control approaches and existing precautionary approaches result in overexploitation and long term declines of fish stocks, as well as reduced economic value, while in contrast, the POMDP approach is able to deliver both ecological and economic recovery despite uncertainty in measurements.  These results have important implications for both global fisheries and ecological management more generally.  In marine fisheries, evidence of such over-exploitation has already been well documented [@Worm2006; @Worm2009; @Costello2016]. Our results suggest that measurement uncertainty may contribute to this pattern, and that reality of measurement errors may frustrate stock recovery under current best practices such as MSY, TAC, and CE.  More generally, the study and application of strategies such as MSY and CE are in no way limited to the decisions in fisheries quotas, but underpin much of conservation and natural resource management in forests and othe terrestrial ecosystems as well [e.g. @Clark2000].  Our results underscore the realization that measurement uncertainty is not negligible.  Neither optimization approaches which ignore uncertainty (like CE) or heuristic approaches to account for that uncertainty (like TAC) are sufficient for this task.  Rather, adaptive approaches such as POMDP, which make use of previous estimates to continually adjust prior beliefs, are required to avoid negative ecological and economic outcomes.  Nevertheless, it may be tempting to see these results as either trivial or impractical, but a closer inspection reveals they are neither.

@@

Third: are the examples considered here too simple?  The strategies we have compared against underpin both current practice and management aspirations.  Together, MSY and TAC-based management reflects much of modern best practice as it is understood in fisheries management and elsewhere. While real-world application of these approaches frequently involves more complex models, the models we have considered here are the very ones under which these strategies were designed and tested, and there is little reason to believe they would perform better under more complex cases such as age-structured models that do not always meet those assumptions [e.g. @Holden2015].  The relatively poor ecological and economic performance of these common best-practices relative to the alternative approach of POMDP is thus reason enough for concern.  This issue is further underscored by the reality of both long-running ecological detioriation [@Worm2006; @Worm2009] and economic costs of global fisheries.

Fourth: is the approach taken here too complex to be feasible? The importance of simple and effective rules of thumb in conservation management has been well documented [e.g. @Chades2011]. Sophisticated and computationally intensive approaches such as POMDP may seem improbable at a time when many areas of natural resource management, even simple, rule-of-thumb based methods struggle to take root.  And yet, in this same age we rely on far more complex and computationally intensive algorithms to perform far more trivial tasks such as determining what advertisements we see.  Surely a planet that can afford such complex optimization for how advertising space is allocated on a small screen can one day manage it's natural resources and conservation efforts with such tools?

<!-- Adaptive management -->

@@

Several limitations that have been studied in fully observed (MDP) optimal
decision problems, such as parameter uncertainty [e.g. @Ludwig1982], model 
uncertainty [@Williams2001; @Boettiger2015] and adaptive management [e.g. @Walters1976]
remain largely open challenges for partially observed systems. Future work 
could extend this analysis to more complex models, such as those with age structure,
as [@Holden2015] does for the fully observed case. Another limiting assumption 
common to MDPs and POMDPs is that of stationary dynamics: that the population 
dynamics equation itself is not changing over time. In reality, forces such as
climate change and other forms of environmental variations violate this
assumption  [@Britten2017]. Direct approaches such as @Fackler2014's
adaptation of Mixed Observability MDP [@Ong2010] do not scale to the number
of states and actions considered here. A value of information (VOI) analysis
for POMDP,such as @Johnson2015 shows for MDP models, could identify when it
is worthwhile to actively reduce measurement error.
